Reversible Regime Change: Climate-driven phytoplankton community shifts in the Cariaco Basin, Venezuela

### 2025-08-14 text-backup.tex
Final Text Revision

>>>>

%OLDCONCLUSION
We found a strong impact of decadal-scale climate variability on phytoplankton community shifts in the tropical coastal ecosystem of the Cariaco Basin. Specifically, phytoplankton cell counts revealed two distinct community clusters, corresponding to periods of high and low upwelling, which aligned with anomalies in the AMO index. We observed no significant differences in overall diversity between the two clusters, except for reduced taxonomic turnover and increased evenness during the low-upwelling phase.
    These community changes did not constitute a permanent regime shift, as a return to the earlier cluster was observed toward the end of the time series (past 2014). Despite multi-annual trends in warming, reduced productivity, and drastic declines in micro-phytoplankton cell counts, the phytoplankton community returned to previous levels once the long-term climatic and local physical conditions were restored. 
    Notwithstanding a number of limitations, our analysis points to the incipient return of the Cariaco Basin to a pre-2004 regime, hinting at an oscillatory nature of phytoplankton communities under global change, an aspect that emphasizes the relevance of long-term time series and deserves further attention.

    The strongest variable predicting changes in the phytoplankton community was the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local conditions through low-frequency natural cycles. We concur with \citeA{nye_ecosystem_2014} that the AMO appears to be a suitable candidate as an indicator of ecosystem state, particularly for the phytoplankton community of the Cariaco Basin. 
    Future studies, ideally based on mathematical modeling, could focus on the relative impact of top-down vs. bottom-up drivers, aspects remaining largely unexplored in the Cariaco Basin.
